Supporting Information

### 2022-03-09 Manuscript/Main_text/Main_text_Oikos.Rmd
read again tomorrow but is looking good

For both floral visitor and plant groups we found different frequencies on motif positions that implied different ecological roles. First, for floral visitors we found that bees had a tendency to appear with higher frequency on motif position that involved generalism while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) showed the opposite trend with higher frequency on motif positions that implied greater level of specialization (**Figure S2 A**). All plant functional groups showed a slight tendency for generalism (**Figure S2 B**). Second, regarding the number of indirect interactions, bees and tall plants with unisexual flowers showed greater frequency on motif positions that had low number of indirect interactions but high of direct interactions (**Figure S3 and Figure S4**). The rest of floral visitor groups showed the opposite tendency and had higher frequencies on motif positions with high number of indirect interactions but low of direct interactions (**Figure S3**). The rest of plant groups did not show major differences on their frequencies on motifs that implied different proportion of indirect-direct interactions (**Figure S4**).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network motifs can capture both direct and indirect interactions, by analysing over- and under-representation of motifs we are able to explore in greater detail ecological interactions [@stouffer2007]. We have explored for the first time the overall structure and representation of network motifs in real plant-pollinator networks and shown that the different motifs up to five nodes are consistently over- and under-represented in comparison with simulated networks indicating the presence of ecological processes shaping their structure. Despite the presence of multiple ecological processes likely operating at the same time (e.g., neutral effects, morphological matching or phenological overlap), we found a tendency for over-represented structures with more densely connected motifs (i.e., complete, fan and asymmetric complete) suggesting that neutral effects are likely the predominant processes governing the structure of these network motifs [@simmons2020]. Thus, as indicated for other network properties like nestedness [@krishna2008; @suweis2013], species relative abundances (i.e., neutral effects) are possibly one of the main determinants of indirect interactions in real plant-pollinator networks.

From a pollinator perspective, we found that generalist network motifs were consistently over-represented (e.g. fan-type motifs; **Figure 2**). For instance, when we compare motifs with equal structure for both trophic levels, we found that floral visitors had always over-representation in the positions that implied the highest number of links. Despite this is in good agreement with the current view of the generalist tendency of pollinators [@waser1996; @olesen2002; @ollerton2017], we found that this over-representation of the generalist network motifs was largely driven by the functional group of bees (e.g., positions 42, 46, 16, 4, 14 and 35 in motifs 15, 17, 7, 2, 6 and 13, respectively; see **Figure 3** for positions and **Figure 4** for motifs) while the rest of the functional groups tended to be under-represented in these generalist motifs and over-represented in motif positions with specialist roles (e.g., positions 11, 21, 28 and 34 in motifs 5, 9, 11 and 13, respectively; see **Figure 3** for positions and **Figure 4** for motifs). This large effect of bees on the overall patterns is no surprise, as bees were the functional group with largest number of interaction records in the whole set of networks as also recorded for other broad scale plant-pollinator network studies [@carvalheiro2014; @ollerton2017]. Unlike floral visitors, plant functional groups showed little differences on its appearance in different motif positions, and thus similar probability of appearing on all of them. If anything, the groups that involved higher levels of outcrossing showed a tendency to be more over-represented and probable on different motif positions. For instance, the group of self-incompatible perennials with large flowers, one of the groups that is expected to have higher levels of outcrossing rate [@goodwillie2010], showed the highest over-representation of all plant functional groups (**Figure 3**); and the group of tall plants with unisexual flowers, the one that is expected to have the highest dependence of pollinators on reproduction [@rodger2021], was the most probable on the generalist motif for plant positions (i.e., plant nodes with 3 to 4 links; **Figure 4**).

The probabilities of appearance of each functional group on the different motif positions were insufficient to predict most of the observed network position combinations (**Figure 5**). This implies that not just the representation of network motifs types but also the identity of the elements of their structure is influenced by ecological processes that determine under- and over-representation of the different position combinations. In addition, we found that just 30$\%$ of the possible 53,250 motif combinations are realized. Thus, similarly to the presence of forbidden links at the network level [@olesen2011; @jordano2016], we also found the existence of forbidden motif positions and those represented almost two thirds of the total posible combinations of indirect interactions. This concurs well with the general idea that plant-pollinator networks have a high proportion of non-realized direct interactions [@jordano2003; @bascompte2013], and as a result, also a high proportion of non-realized combinations of indirect interactions. Some of this non-observed interactions are probably due to under-sampling [@dorado2011; @jordano2016]. However, a large fraction of the possible interactions do not take place because of life-history constraints such as species co-occurrence, trait matching processes and phenological uncoupling [@olesen2011; @bartomeus2016; @jordano2016]. Despite the advances in the understanding on the ecological processes that influence species pairwise interactions [@bartomeus2016; @peralta2020], we still know little of how these processes determine competitive and facilitative interactions in mutualistic networks. Thus, as suggested by @simmons2020, exploring the proportion of these realized network motifs that are competitive or facilitative is a necessary next step to improve our understanding of the architecture of biodiversity [@bascompte2007].

@@

>>>>

For both floral visitor and plant groups we found different frequencies on motif positions that implied different ecological roles. First, for floral visitors we found that bees had a tendency to appear with higher frequency on motif position that involved generalism while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) showed the opposite trend with higher frequency on motif positions that implied greater level of specialization (**Figure S2 A**). All plant functional groups showed a slight tendency for generalism (**Figure S2 B**). Second, regarding the number of indirect interactions, bees and tall plants with unisexual flowers showed greater frequency on motif positions that had low number of indirect interactions but high number of direct interactions (**Figure S3 and Figure S4**). The rest of floral visitor groups showed the opposite tendency and had higher frequencies on motif positions with high number of indirect interactions but low of direct interactions (**Figure S3**). The rest of plant groups did not show major differences on their frequencies on motifs that implied different proportion of indirect-direct interactions (**Figure S4**).

@@

Understanding the global patterns of network structure is key to unravel the processes that govern community dynamics [@bascompte2007; @bastolla2009; @guimaraes2020]. However, common analytical approaches are unable to capture the full complexity of ecological systems [@simmons2019; @bergamo2021]. Because network motifs can capture both direct and indirect interactions, by analysing over- and under-representation of motifs we are able to explore in greater detail ecological interactions [@stouffer2007]. We have explored for the first time the overall structure and representation of network motifs in real plant-pollinator networks and shown that the different motifs up to five nodes are consistently over- and under-represented in comparison with simulated networks indicating the presence of ecological processes shaping their structure. Despite the presence of multiple ecological processes likely operating at the same time (e.g., neutral effects, morphological matching or phenological overlap), we found a tendency for over-represented structures with more densely connected motifs (i.e., complete, fan and asymmetric complete). As suggested in @simmons2020, a higher proportion of more densely connected motifs could reflect that neutral effects are the predominant ecological process governing the structure of these network motifs. Thus, as indicated for other network properties like nestedness [@krishna2008; @suweis2013], species relative abundances (i.e., neutral effects) are possibly one of the main determinants of indirect interactions in real plant-pollinator networks.

Overall, generalist network motifs for floral visitors were consistently over-represented (e.g. fan-type motifs; **Figure 2**). For instance, when we compare motifs with equal structure for both trophic levels, we found that floral visitors had always over-representation in the positions that implied the highest number of direct links. Despite this is in good agreement with the current view of the generalist tendency of pollinators [@waser1996; @olesen2002; @ollerton2017], we found that this over-representation of the generalist network motifs was only driven by the taxonomic group of bees (**Figure S2**) while the rest of the taxonomic groups tended to be under-represented in these motif that involved generalism and over-represented in motif positions with specialist roles. This large effect of bees on the overall patterns is no surprise, as bees were the group with largest number of interaction records in the whole set of networks as also recorded for other broad scale plant-pollinator network studies [@carvalheiro2014; @ollerton2017]. Interestingly, bees also were the group with lower number of indirect interactions while the rest of floral visitor groups showed clearly higher number of indirect interactions indicating that these groups may experience higher levels of competitive interactions due to greater overlap of floral resources [@thomson2020]. Unlike floral visitors, plant functional groups showed little differences on its appearance in different motif positions, and thus similar probability of appearing on all of them. However, the groups that involved higher levels of outcrossing showed a tendency to be more over-represented and probable on the different motif positions. Remarkably, the group of tall plants with unisexual flowers, the one that is expected to have the highest dependence of pollinators on reproduction [@rodger2021], showed the lowest level of indirect interactions but the highest frequency of records for all groups which involves a high number of direct interactions that are likely to minimise negative impacts of pollinator sharing (e.g., heterospecific pollen arrival).

The probabilities of appearance of each functional group on the different motif positions were insufficient to predict most of the observed network position combinations (**Figure 5**). This implies that not just the representation of network motifs types but also the identity of the elements of their structure is non-random. Notably, we found that just 30$\%$ of the possible 53,250 motif combinations are realized. Thus, similarly to the presence of forbidden links at the network level [@olesen2011; @jordano2016], we also found the existence of forbidden motif positions and those represented almost two thirds of the total possible combinations of indirect interactions. This concurs well with the general idea that plant-pollinator networks have a high proportion of non-realized direct interactions [@jordano2003; @bascompte2013], and as a result, also a high proportion of non-realized combinations of indirect interactions. Some of this non-observed interactions are probably due to under-sampling [@dorado2011; @jordano2016]. However, a large fraction of the possible interactions do not take place because of life-history constraints such as species co-occurrence, trait matching processes and phenological uncoupling [@olesen2011; @bartomeus2016; @jordano2016]. Despite the advances in the understanding on the ecological processes that influence species pairwise interactions [@bartomeus2016; @peralta2020], we still know little of how these processes determine competitive and facilitative interactions in mutualistic networks. Thus, as suggested by @simmons2020, exploring the proportion of these realized network motifs that are competitive or facilitative is a necessary next step to improve our understanding of the architecture of biodiversity [@bascompte2007].

@@

**REWRITE**

### 2022-03-09 Manuscript/Supplementary_Material/Supporting_information_Oikos.Rmd
read again tomorrow but is looking good

**Figure S2** Over and under-estimated functional groups on motif positions.

**Figure S3** Plant trait composition of functional groups.

**Figure S4** Plant dendrogram with functional groups.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.align = 'center', fig.cap="\\textbf{Figure S2.}"}

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,out.width="90%",fig.align="center", fig.width=10,fig.height=12, fig.cap="\\textbf{Figure S3.}"}

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,out.width="90%",fig.align="center", fig.width=10,fig.height=12, fig.cap="\\textbf{Figure S4.}"}

>>>>

**Figure S2** Association between the frequency and specificity of the different groups on motif positions.

**Figure S3** Frequency of floral visitor groups and proportion of indirect-direct interactions.

**Figure S4** Frequency of plant groups per motif and proportion of indirect-direct interactions.

**Figure S5** Over and under-estimated groups on motif positions.

**Figure S6** Plant trait composition of functional groups.

**Figure S7** Dendrogram with plant functional groups.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.align = 'center', fig.cap="\\textbf{Figure S2.} Association between the frequency and specificity on the different motif positions for floral visitor (panel A) and plant (panel B) groups. The different numerical categories of the plant groups correspond to: 1 (selfing herbs), 2 (small outcrossing perennials), 3 (self-incompatible perennials with large flowers), 4 (tall plants with small unisexual flowers) and 5 (short-lived outcrossers with long zygomorphic flowers)."}

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,out.width="90%",fig.align="center", fig.width=10,fig.height=12, fig.cap="\\textbf{Figure S3.} Frequency of the different floral visitor groups on the different motif positions aggregated by the proportion of indirect-direct interactions for the different motif positions."}

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,out.width="90%",fig.align="center", fig.width=10,fig.height=12, fig.cap="\\textbf{Figure S4.} Frequency of the different plant groups on the different motif positions aggregated by the proportion of indirect-direct interactions for the different motif positions."}
